---
title: "Agricultural Survey of African Farm Households"
author: "Denis Mwaniki"
date: " `r as.Date(Sys.time())` "
output:
  pdf_document:
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE, cache=F}
knitr::opts_chunk$set(echo = F, warning = F, message = F, fig.width = 10, fig.height = 10)

```

## Importing Datasets and Preprocesing

### Importing Libraries

```{r packages}

### clear environment
rm(list = ls())



# load libraries

library(tidyverse)
library(dplyr)
library(readxl)
library(ggplot2)
library(scales)

```

### Reading the data

```{r datasets, echo=TRUE}

### check working directory
getwd()

setwd("D:/Njambanene/Njambanene/R/R/Case Study/FACS")
### read dataset

dt0 <- readxl::read_xlsx("../FACS/Case_2_Statistics.xlsx")
### preview
glimpse(dt0)
head(dt0)

```


### Cleaning

```{r dataset cleaning1, echo=TRUE}

#### mapping gender labels

dt1 <-dt0 %>%
  mutate(across(starts_with("gender"),~ recode(as.character(.x)  ,'1' = "Male",'2' = "Female",.missing = "Female")) )

head(dt1)

```


```{r dataset cleaning2, echo=TRUE}

#### mapping married labels

dt2 <- dt1 %>%
  mutate(married1 = recode(as.character(married1),'1' = "Married",'2' = "Never Married",'3' = "Previously Married",'4' = "NA"))

head(dt2)

```

## Exploratory Data Analysis

```{r hh size, echo=TRUE}

#### get size of hh
##### men & women in hh

dt3 <- dt2 %>%
  rowwise() %>%
  mutate(
    num_men = sum(c_across(starts_with("gender")) =="Male",na.rm = T),
    num_women = sum(c_across(starts_with("gender")) =="Female",na.rm = T),
    hh_size = num_men + num_women
  ) %>%
ungroup()

head(dt3)

```


```{r female hh, echo=TRUE}

#### print female in hh

dt4 <- dt3 %>%
  rowwise() %>%
  mutate(
    female_hh = ifelse(num_women>=5,1,0)) %>%
  ungroup()

head(dt4)

```

### Validation Checks

```{r na check1, echo=TRUE}

### check NAs
#### total NAs in data
sum(is.na(dt4))

```


```{r na check2, echo=TRUE}

#### atleast a row with NA 
sum(rowSums(is.na(dt4))>0)

```

```{r na check3, echo=TRUE}

#### count Nas per col
colSums(is.na(dt4))

```


```{r na imputation, echo=TRUE}

#### replace NA in a col w 0
dt4$incfarm[is.na(dt4$incfarm)] <- 0


```


```{r outlier check, echo=TRUE}

### check for outliers

inc_summary <- dt4 %>%
  filter(incfarm > 0 ) %>%
  summarise(
    mean_income = mean(incfarm),
    min_income = min(incfarm),
    q1 = quantile(incfarm,0.25),
    median_income = median(incfarm),
    q3 = quantile(incfarm,0.75),
    max_income = max(incfarm)
  )

print(inc_summary)


```

## Analysis

### Overall Breakouts

```{r boxplot, echo=TRUE}

#### Plot a boxplot of incfarm

ggplot(dt4, aes(y = incfarm)) +
  geom_boxplot(fill = "lightblue")+
  scale_y_log10(
    labels = scales::comma_format(),
    breaks = scales::log_breaks(n = 10)
  )+
  labs(title = "Farm Income Distribution", y= "Income")


```



```{r pop by country, echo=TRUE}

#### number of people per country

country_pop <- dt4 %>%
  select(hh_size,country) %>% 
  group_by(country) %>%
  summarise(total_pop = sum(hh_size)) %>%
  mutate(percentage = total_pop/sum(total_pop)*100)


```


```{r plot 1, echo=TRUE}

##### plot 1

ggplot(country_pop,aes(x = country,y= total_pop, fill = country)) +
  geom_bar(stat = "identity",width = 0.7) +
  geom_text(aes(label = paste0(round(percentage,1),"%")),
            vjust = -0.3,size = 3.5) +
  geom_text(aes(label = total_pop),
            vjust = 1.2, size = 3.5) +
  labs(title = "Number of people per country",x = "Country",y= "Total Population") +
  theme_minimal()+
  theme(legend.position = "none")


```


```{r prop of female hh, echo=TRUE}

#### prop of female hh

proportion_of_fhh <- mean(dt4$female_hh) * 100

print (proportion_of_fhh)


```

```{r plot 2, echo=TRUE}

#### gender distr
gender_dt <- data.frame(
  category = c("Female-headed","Male-headed"),
  counts = c(sum(dt4$female_hh),nrow(dt4) - sum(dt4$female_hh))
)

##### plot 2
ggplot(gender_dt,aes(x = "",y = counts, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(round(counts/sum(counts) * 100,2),"%")),
            position = position_stack(vjust = 0.5))+
  labs(title = paste0("Proportion of female households: ",
                      round(proportion_of_fhh,2), "%")) +
  theme_void()+
  theme(legend.title = element_blank(),
        legend.position = "bottom")


```


### Distribution of farm income in Ghana

```{r ghana farm income, echo=TRUE}

### filter for ghana
gh_f_inc <- dt4 %>%
  filter(country == "ghana") 

### print summary stats
gh_f_inc_summary <- gh_f_inc %>%
  select(country,incfarm) %>%
  summarise(
    mean_income = mean(incfarm),
    median_income = median(incfarm),
    sd_income = sd(incfarm),
    min_income = min(incfarm),
    max_income = max(incfarm)
  )

print(gh_f_inc_summary)

```



```{r plot 3, echo=TRUE}

##### plot 3

ggplot(gh_f_inc,aes(y = incfarm))+
  geom_boxplot(fill = "lightblue") +
  scale_y_log10(
    labels = scales::comma_format(),
    breaks = scales::log_breaks(n = 10)
  )+
  labs(title = "Farm Income Distribution Plot",y = "Income")

```

### Age Distribution

```{r plot 4, echo=TRUE}

##### plot 4
ggplot(dt4,aes(x = age1)) +
  geom_histogram(binwidth = 5,fill = "blue", color = "black") +
  labs(title = "Age Distribution", x = "Age", y = "Frequency")

```


```{r gender stats, echo=TRUE}

#### gender descriptive stats
gender_analysis <-   gh_f_inc %>% 
  group_by(female_hh) %>%
  summarise(
    mean_income = mean(incfarm),
    sd_income = sd(incfarm)
  )

#### gender
t.test(incfarm ~female_hh, data = gh_f_inc)

```

#### Overall Conclusion:

* No Significant Difference: The t-test results indicate that there is no statistically significant difference in mean income between group 0 and group 1. The difference observed in the sample means could likely be due to random chance.
* Fail to Reject the Null Hypothesis: The high p-value and the confidence interval including 0 mean that you do not have enough evidence to reject the null hypothesis of no difference in means between the two groups.
* This analysis suggests that, based on the data provided, the incomes in the two groups are not significantly different.

### Education Distribution

```{r education, echo=TRUE}

#### education 
aov_result <- aov(incfarm ~educ1, data = gh_f_inc)
summary(aov_result)

```


#### Overall Interpretation:

* Statistical Significance: The factor educ1 has a statistically significant effect on the response variable (e.g., income) at the 5% significance level. This means that the difference in the response variable across the levels of educ1 is unlikely to be due to random chance.
* Magnitude of Effect: Although the result is statistically significant, the F value is relatively modest (5.585), suggesting that while educ1 has an effect, it might not explain a large proportion of the variance in the response variable.
* In summary, the ANOVA results indicate that the variable educ1 significantly affects the response variable, suggesting that differences in educ1 levels are associated with differences in the response variable (e.g., income).

### Age Distribution

```{r age, echo=TRUE}

#### age
aov_result1 <- aov(incfarm~age1, data = gh_f_inc)
summary(aov_result1)

```


#### Overall Interpretation:

* Statistical Significance: The factor age1 has a statistically significant effect on the response variable (e.g., income) at the 5% significance level. This means that the difference in the response variable across the levels of age is unlikely to be due to random chance.
* Magnitude of Effect: Although the result is statistically significant, the F value is fairly high (13.45), suggesting that age1 has an effect, and it may explain a large proportion of the variance in the response variable.
* In summary, the ANOVA results indicate that the variable age1 significantly affects the response variable, suggesting that differences in age1 levels are associated with differences in the response variable (e.g., income).

### Sickdays Distribution

```{r sickdays, echo=TRUE}

#### sickdays
aov_result2 <- aov(incfarm~sickdays , data = gh_f_inc)
summary(aov_result2)

```



#### Overall Interpretation:

* No Significant Difference: The factor sickdays has no statistically significant effect on the response variable (e.g., income) at the 5% significance level. This means that the difference in the response variable across the levels of sickdays could likely be due to random chance.
* Magnitude of Effect: While the result is not statistically significant, the F value is almost 0 (0.309), suggesting that sickdays has minimal effect, and it cannot explain the variance in the response variable.
* In summary, the ANOVA results indicate that the variable sickdays does not significantly affect the response variable, suggesting that differences in sickdays levels are not associated with differences in the response variable (e.g., income).


```{r multiple linear regression, echo=TRUE}

#### multiple linear regression
lm_result <- lm(incfarm ~ female_hh + educ1 + age1 + sickdays ,data = gh_f_inc)
summary(lm_result)

```


